---
title: "R - Postgres Tutorial"
output: html_notebook
---
#### Install the package

You can manipulate and query a [PostgreSQL](https://www.postgresql.org/) database from within R. The Package *RPostgreSQL* offers a high level binding and some helpful querying functions for handling a PostgreSQL instance. This database system is especially helpful in case you want to handle geodata as it includes a full featured GIS system in the extension [PostGIS](http://postgis.net/).
At the same time, PostgreSQL can be tuned to be a high performance database system, which makes it extemely help- and powerful for hydrologists as we have both: geodata and huge (complex) data amounts. Last but not least, PostGIS is working very well with QGis, GRASS or SAGA GIS and some GIS Web Servers like Geoserver or Mapserver. This makes the R / Python - PostgreSQL - Online/Offline GIS stack maybe the most powerful software stack for (geo)data science.<br>
First, install and load the library.

```{r}
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
require(RPostgreSQL)
```

### Connection 

RPostgreSQL is build like most other database drivers, across languages. This means in most cases it does not matter which kind of RDBMS you want to connect to and from which language. Connecting to PostgreSQL from R works a lot like connecting SQLite from Pyhton or MySQL from C++. First, you need a driver. This class instance defines the database system you want to talk to and is meant to be a layer between your abstract R classes and the RDBMS specific SQL code. This driver can then be used to build a connection object to a specific database instance (like the one we will be using). Changing the connection to another instance will enable you to run the exact same code on more than one machine. After you are finished, the connection has to be closed again, in order to free these resources on the server. ddd

```{r}
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='openhydro.de', port=5432, user=.rs.askForPassword('Provide the user'), 
                 password=.rs.askForPassword('Provide the password'), dbname='dwd')

```

From here, RPostgreSQL offers different function for communicating with the PostgreSQL instance behind that connection. These functions reach from rather low level SQL query execution to high level get-a-whole-table-and-turn-to-data.frame-functions. The most important interfaces are:
<ul>
  <li>**dbGetQuery(_connection_, _SQL_):** send the query and fetch the result</li>
  <li>**dbExecute(_connection_, _SQL_):** send the query without fetch</li>
  <li>**dbReadTable(_connection_, _table.name_)** load the whole table into a data.frame</li>
  <li>**dbWriteTable(_connection_, _table.name_)** write the data.frame into the specified table. Caution the structure must match exactly</li>
  <li>**dbListTables(_connection_)** list all tables in the connected database</li>
<ul>

### Working with data

As a first exercise, we want to read the whole station list from the database.
```{r}
dbListTables(con)
```

Then load the stations
```{r}
stations <- dbReadTable(con, 'stations')
stations
```

We will head to the unrecognized column later on. At first, let's add a arbitrary station and remove it again. Remember, that you have to preserve the table structure. The most easiest approach is therefore to append a new row to the just created data frame. Ignore the last value for now. Lastly, it is important to set the 'append' flag, in order to append the data to the end of the table.

```{r}
import <- data.frame(id=9999, name='Mirkos Station', short_stat='MIRK', elev=2500, geometry="POINT(7.8 48.0)")
names(import) <- names(stations)
dbWriteTable(con, 'stations', import, append=TRUE, row.names=FALSE)
```

### Executing SQL

In order to check, wether our data reached the database, we will create our own custom SQL query to ask the database for exactly that row. We will go into more details of the SQL language in the Introductory SQL notebook. That means you do not have to understand each part of the query at this moment. Concentrate more on the technique of sending queries. As we are asking for data, we cannot use the *dbExecute* method, as this does not fetch possible results. Therefore we have to use the *dbGetQuery* function here, as this one will fetch the results.<br>

<div class="alert alert-warning">Under the hood fetching the result from the connection is a multistep process. First, the SQL has to be executed in the database. Then, the connection has to be flushed. Next, the response from the DB server is cached and is finally converted from a raw bytes objects to a R object.</div>

```{r}
sql <- 'Select * from stations where id=9999'
dbGetQuery(con, sql)
```

There are two interesting things to notice here:
<ol>
  <li>The geometry column got somehow converted. It was stated that you do not have to understand this input, but the returned values are definitely different.</li>
  <li>Executing SQL is quite straitforward. We are now able to come up with literally any manipulation, filter or selection we wish.</li>
</ol>
RStudio Notebooks are also a great way of working with databases. Beside a code chunk in R, it can also handle SQL chunks. All we need
is the dbConnection object referenced in the 'con' variable. All we need to do, is placing the SQL directly into the chunk and RStudio will do the rest for us. <br>
Although this makes coding on databases easy and improves the readability of the notebook, be aware of the actual steps happing here in case you ever have to build a script without RStudio. Do not get too dependend on a specific IDE, even if it's a great IDE.<br>
However, the code from the last chunk could also be run like:

```{sql connection=con}
select * from stations where id=9999
```

This leads to exactly the same result, while we can concentrate on the SQL language. Lastly, we will remove the created line again as this station does not really exist. The <span style='color:blue'>DELETE</span> statement follows the same syntax as <span style='color:blue'>SELECT</span>.<br>
<div class="alert alert-danger"><span class="glyphicon glyphicon-exclamation-sign" aria-hidden="True"></span> **CAUTION**: Be absolutely sure about your statement as SQL does not ask if you really want to delete data. It will just be gone. Without a WHERE clause, the whole table will be deleted!</div>

```{sql connection=con}
delete from stations where id=9999
```

### Closing the connection

Do not forget to ALWAYS close the connection. The PostgreSQL Server will limit the number of connections. 
Moreover, each connection uses some of the system resources and too many connections will unnecessarily slow down the system.

```{r}
dbDisconnect(con)
```