---
title: "Advanced SQL techniques"
author: "Mirko Mälicke"
output:
  html_document: 
    toc: yes
    theme: flatly
  html_notebook:
    toc: yes
    theme: flatly
  pdf_document:
    toc: yes
---

```{r, echo=FALSE}
# require the package
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
if (!("getPass" %in% installed.packages())){
  install.packages("getPass")
}
require(RPostgreSQL)
require(getPass)

# establish the connection
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='openhydro.de', port=5432, user=getPass('Provide the user'), 
                 password=getPass('Provide the password'), dbname='dwd')
```

### Temporarily creating objects

Especially for development and data analysis tasks it is very useful to create temporary results. This saves a lot of time and will keep
 your database clean as you do not have to remember which objects were only intermediate and can be droped.
The SQL language knows the <span style="color:blue">TEMPORARY</span> keyword, which can be used along with <span style="color:blue">CREATE</span> statements. 
This is most often used on the creation of tables and views. The temporary tables can be used just like persistent tables, but once you close the 
connection used for creation, the table will automatically be droped. 
This can save you from a lot of cleanup work. 

<div class="alert alert-warning">Some SQL clients open and close a connection on each command issued in order to keep the number of connection small and
prevent the user from keeping open connections. You will have to </div>



### cleanup

```{r}
dbDisconnect(con)
```

