---
title: "R Notebook"
author: "Mirko Mälicke"
output: html_notebook
---

<strong>This is not a mandatory exercise. You can do this exercise in case you have still trouble with basic SQL query syntax and logic.</strong><hr>
In case you need more practice with basic SQL query structure and the logic behind a data manipulation using SQL, this exercise is perfect. 
It does not include any environmental data, thus you will be forced to think outside your comfort zone of time-series data. You will only
need the `nfl` table in the database. This collects all NFL (National Football League) games of the 2016 / 2017 regular season (no Playoffs included). 
Each entry gives the data, home team, vistor team and the score of both teams. Try to solve all given tasks.

```{r setup, echo=FALSE}
require(RSQLite)
drv <- dbDriver('SQLite')
con <- dbConnect(drv, '../utils/exercises.sqlite')
```



### Get to know the data

How many games does a regualar season in the NFL include?
```{sql connection=con}

```

How many teams are organized in the NFL?
```{sql connection=con}

```

How many games does a NFL team complete? Will all teams play against each other (just like in the Bundesliga)?
```{sql connection=con}

```

Which are the NFL teams of 2016?
```{sql connection=con}

```

One of my favorite Quarterbacks, Cam Newton, played 2016 on my birthday, which is Sep. 8th. Did Cam win his game?
```{sql connection=con}

```

How many teams from New York play in the NFL?
```{sql connection=con}

```

How many Los Angeles teams played in the NFL in 2016?
```{sql connection=con}

```

In 2017 preseason the Los Angeles Chargers celebrated a close victory over the Los Angeles Rams of 21:19. As you learned from the previos chunk, both teams relocated from somewhere else to Los Angeles. Where were the Rams and Chargers located until 2016?
```{sql connection=con}

```

### Working with the data

Enough preliminary data exploration. Now, let's work with the data. <br>
Create a view on the nfl table which adds a boolean information on whether the home team won the game or not.
```{sql connection=con}

```

Is there a home game advantage?
```{sql connection=con}

```

Which team had the most home victories, which won the fewest? Select both teams in a union statement.
```{sql connection=con}

```

Because of the American Football rules ties are very sparse in the NFL. From 1997 to 2015 there were overall only three ties. How many ties happened 2016?
```{sql connection=con}

```

### Aggregation

Other than in the Bundesliga, a NFL team won't earn points for a victory. The teams are split up into _divisions_ of four teams. The team with most victories will win the division and go to the playoffs. Therefore the record is the most important statistic for the teams.
Before the 11th game day, the Green Bay Packers had a `Green Bay Packers 4-6` record, meaning they lost 6 out of 10 games.<br>
<br>
How did the Green Bay Packer do? What was their record at the end of the regular season?
```{sql connection=con}

```

Calculate the closing recordfor all teams and persist the result as a new view.
```{sql connection=con}

```


Tom Brady, the Patriot's Quarterback, is for many people the best Quarterback of all times. Because of the _Deflategate_ in the Superbowl 2015/2016 he was suspended for four games. How did the team perform with him, how without him?
```{sql connection=}

```

In the NFL all are about statistics. You will find a statistic for literally everything. Even the salary of players is highly depended on their stastistical performance. The most important offensive statistic is the point/game value. The team with the highest value is considered to be the best offensive team. Which one was that?
```{sql connection=con}

```

The opposite is used for the defense. The less points the opponent scores per game, the better the defense is. Which one was the best defensive team?
```{sql connection=con}

```

Calculate the total points in home and visiting games for each team. Use the ratio of both to rank all teams. 
```{sql connection=con}

```

Calculate the offensive score (points / game) for all teams grouped by home and visiting. Do the same thing for the defensive score. 
Use these two datasets to group the offensive/defensive scores by the amount of won games and save them to R data.frames. Use R to
visualize the correlation between the number of games won and the four different scores. Can you find any correlations?
```{sql connection=con, output.var="off.home"}
-- query the offensive home score here

```
```{sql connection=con, output.var="off.visit"}
-- query the offensive visiting score here

```
```{sql connection=con, output.var="def.home"}
-- query the defensive home score here

```
```{sql connection=con, output.var="def.visit"}
-- query the defensive visiting score here

```
```{r}
# correlation analysis
# off.home, off.visit, def.home and def.visit are available here

```



### cleanup
```{r, echo=FALSE}
dbDisconnect(con)
```

