---
title: "SQL Exercises"
output: html_notebook
---

This notebook collects some basic SQL exercises for the sql_introduction.Rmd lecture. You can solve them and create a pull request for your soultion in the master Github branch.

```{r, include=FALSE}
# require the package
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
require(RPostgreSQL)

# establish the connection
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='openhydro.de', port=5432, user=.rs.askForPassword('Provide the user'), 
                 password=.rs.askForPassword('Provide the password'), dbname='dwd')
```


### SELECT

Select all stations located lower than 400 meter.
```{sql connection=con}

```

How many stations start with the letter 'W'?
```{sql connection=con}

```

Are there more stations starting with 'M' or 'W'?
```{sql connection=con}

```

Are the 'M' or the 'W' stations higher in mean?
```{sql connection=con}

```

### SELECT & Grouping

Find the 5 starting letters in station names with the most occurences.
```{sql connection=con}

```

Find the 10 starting letters in station names with the lowest mean elevation.
```{sql connection=con}

```

 


### cleanup
```{r, include=FALSE}
dbDisconnect(con)
```

