---
title: "PostGIS"
author: "Mirko Mälicke"
output:
  html_document: 
    toc: yes
    theme: flatly
  html_notebook:
    toc: yes
    theme: flatly
  pdf_document:
    toc: yes
---

```{r, echo=FALSE, include=T}
# require the package
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
if (!("getPass" %in% installed.packages())){
  install.packages("getPass")
}
require(RPostgreSQL)
require(getPass)
require(ggplot2)

# establish the connection
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='openhydro.de', port=5432, user=getPass('Provide the user'), 
                 password=getPass('Provide the password'), dbname='dwd')
```

Select the biggest catchment from the einzugsgebiete table.
```{sql connection=con}

```

Select the biggest catchment from the einzugsgebiete table, whichs' name starts with an 'S'.
```{sql connection=con}

```

Select the smallest catchment fron the einzugsgebiete table, which contains a station.
```{sql connection=con}

```

Select the two stations which are closest / farest away from your own location in a single <span style="color:blue">UNION</span> statement.
```{sql connection=con}

```

In a new project, you would like to build a new weather station either close to the Herderbau or near the Münster in Freiburg. You decide to chose the 
location which has less other stations in a 1.5 km radius
```{sql connection=con}

```

In class, you joined the vegetation class table with its lookup table and calculated the areas for each feature. As you might have noticed, there are duplicates in the description field. 
This is obviously the reason why a lookup table was used. Aggregate these classes together and calculate the overall area for each cover class.
```{sql connection=con}

```

Redo the last step but query the data with nice looking values like '34.1 km²' for the area and add a ratio field, which should give the share of each class on the total area in a '20.1%' style.
```{sql connection=con}

```

Do the einzugsgebiete and vegetation_cover tables cover the same area?
```{sql connection=con}

```




### cleanup

```{r}
dbDisconnect(con)
```

