---
title: "Advanced SQL"
author: "Mirko Mälicke"
output:
  html_notebook:
    theme: flatly
    toc: yes
  html_document:
    theme: flatly
    toc: yes
  pdf_document:
    toc: yes
---

```{r, echo=FALSE, include=T}
# require the package
if (!("RPostgreSQL" %in% installed.packages())){
  install.packages("RPostgreSQL")
}
if (!("getPass" %in% installed.packages())){
  install.packages("getPass")
}
require(RPostgreSQL)
require(getPass)
require(ggplot2)

# establish the connection
drv <- dbDriver('PostgreSQL')
con <- dbConnect(drv, host='openhydro.de', port=5432, user=getPass('Provide the user'), 
                 password=getPass('Provide the password'), dbname='dwd')
```

Build a temporary table called *name_precipitation*, where you should replace *name* with your actual name. This table shall contain 
all measurements from the data table holding precipitation information. 
```{sql connection=con}

```

Create a view on this table that holds monthly statistics for every station id. Calculate the monthly sum, the highest observed daily 
rainfall sum, the number of extreme rainfalls (daily precipitation > 30).
```{sql connection=con}

```

Create a view on the last view, that aggregates the same statistics for anual overviews.
```{sql connection=con}

```

Which station observed the highest monthly rainfall sum?
```{sql connection=con}

```

Which station observed the most extreme rainfall occurences within one month? Which within one year?
```{sql connection=con}

```

Which was the year with the most overall extreme rainfall occurences? (ignoring the station)
```{sql connection=con}

```

Which is the station with the most extreme rainfall events within this year (from the last chunk)? 
Which one with the most extreme rainfall events overall?
```{sql connection=con}

```

Create a view, that includes only the records from _data_ which fulfilled the extreme rainfall constraint.
Join another column to this view, that holds the _type of precipitation_ variable. From this Join, create a histogram of _type of precipitation_. Calculate an overall histogram.
```{sql connection=con}

```

**OPTIONAL:** Can you redo the last step and do the same analysis for 'per station' histograms?
```{sql connection=con}

```

Use the insights of the last two chunks and write two sentences about the _type of precipitation_ variable. How useful is it? 
<div class="alert alert-info">
Answer:

</div>

Load the anual overview on extreme rainfalls into a data.frame and plot a violine plot of the amount of extreme rainfalls per station.
```{sql connection=con, output.var="rain.annual"}

```
```{r}
View(rain.annual)

```

Redo the last exercise for the monthly extreme rainfall counts. Does this change the picture?
What can you say about the distribution of extreme rainfall events from these plots?
```{sql connection=con, output.var="rain.monthly"}

```
```{r}
View(rain.monthly)

```
<div class="alert alert-info">
Answer:

</div>

Load the locations for the used stations from the database.
```{sql connection=con, output.var="locations"}

```
```{r}
View(locations)
```


Plot these locations and represent any measure for annual variance in extreme rainfall events by the point size of the used locations.<br>
**OPTIONAL:** Additionally mark the station showing the most and least recorded extreme rainfalls by another marker into the same plot. (eg. use a green and a red cross or star.)
```{r}

```

How can you interpret this result?
<div class="alert alert-info">
Answer:

</div>







